Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2023-05-06 doc/manuscript/ESSDmanuscript.Rmd
check/update through end of section 3.3

**Abstract.**
Forests are critical for climate change mitigation and constitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhoouse gas emissions and removals.
We begin by reconciling terminology and mapping ForC fields into EFDB.
This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
At the time of writing, ForC contained ~`r n_ind_records_EFDB_variables` independent records that would be relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
ForC represents a diversity of stand ages, although records for young stands are primarily limited to C stocks, as opposed to net increments of fluxes.
This distribution of records is generally reflected in the subset of records that have been submitted to EFDB to date.
In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C accounting.

@@

The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
For C currently contains `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
As such, ForC is positioned to improve forest C accounting through the transfer of data to EFDB.
The purpose of this publication is to document that process and provide recommendations for future improvements.

@@

We added a total of `r n_new_variables` new EFDB-relevant variables to the set of named and defined variables (Fig. `r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one. 
This included eleven increment variables, adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
Further, we added variables capturing the belowground component of woody mortality (*woody.mortality_root*) and the combined aboveground and belowground components of woody mortality (*woody.mortality*).
Although most of these variables currently lack records in ForC, the structure exists such that records can be populated over time. 
Finally, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including OL).

@@

## Quality control measures
Prior to releasing ForC v4.0, we executed several quality control measures.
First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_using_inreview] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision *(Issue #29)[https://github.com/forc-db/ForC/issues/229]*.
Third, to identify erroneous climate data... *(Issue #212)[https://github.com/forc-db/ForC/issues/212]*.

Because ForC v4.0 contained known duplicate records, we used R scripts to remove likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
Henceforth, we refer to the set of records with likely duplicates removed as "independent records".
All records sent to EFCB were ensured to be original through manual review, as detailed below.

>>>>

**Abstract.**
Forests are critical for climate change mitigation and constitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhoouse gas emissions and removals.
We begin by reconciling terminology and mapping ForC fields into EFDB.
This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records that would be relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
ForC represents a diversity of stand ages, although records for young stands are primarily limited to C stocks, as opposed to net increments of fluxes.
This distribution of records is generally reflected in the subset of records that have been submitted to EFDB to date.
In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C accounting.

@@

The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
As such, ForC is positioned to improve forest C accounting through the transfer of data to EFDB.
The purpose of this publication is to document that process and provide recommendations for future improvements.

@@

We added a total of `r n_new_variables` new EFDB-relevant variables to the set of named and defined variables (Fig. `r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one. 
The majority of these were increment variables (n=11), adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
Further, we added variables capturing the belowground component of woody mortality (*woody.mortality_root*) and the combined aboveground and belowground components of woody mortality (*woody.mortality*).
Although most of these variables lacked records in ForC as of `r format(Sys.time(), '%B %d, %Y')`, their addition gave the structure such that records can be populated over time. 
Finally, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including OL).

@@

## Quality control measures
Prior to releasing ForC v4.0, we executed several quality control measures.
First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_implementing_2022] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision.
Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version **##** <!--need this from Becky--> [@climwin_ref] based on site coordinates. 
Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or transfer to EFDB.

Because ForC v4.0 contained known duplicate records, we used R scripts to identify likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
Henceforth, we refer to the set of records with likely duplicates removed as "independent records".
All records sent to EFDB were ensured to be independent and original through manual review, as detailed below.

### 2023-05-06 doc/manuscript/ESSDmanuscript.tex
check/update through end of section 3.3

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned emissions reductions
under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accesible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
accounting if made available through the Emission Factor Database (EFDB)
of the International Panel on Climate Change (IPCC). Here, we develop
and document a process for semi-automated transfer of data from ForC
into the EFDB, assess the data available and transferred to date, and
provide recommendations for improving forest data collection, analysis,
and reporting to improve accounting of forest-sector greenhoouse gas
emissions and removals. We begin by reconciling terminology and mapping
ForC fields into EFDB. This process required some updates to the ForC
database structure, leading to the release of a new version of ForC
(v4.0; described here). At the time of writing, ForC contained
\textasciitilde17204 independent records that would be relevant to EFDB,
1214 of which have been submitted to date. Among the data in ForC, there
is disproportionate representation of biomass (particularly aboveground)
stocks, with far fewer records for dead organic matter and soil C, and
relatively few or no records for net annual increments or C fluxes into
(gains) or out of (losses) the IPCC-defined C pools. Geographic
representation is also quite uneven, with the highest densities of
relevant records in temperate forests, and with relatively scant
representation of tropical forests in Africa and Asia. ForC represents a
diversity of stand ages, although records for young stands are primarily
limited to C stocks, as opposed to net increments of fluxes. This
distribution of records is generally reflected in the subset of records
that have been submitted to EFDB to date. In the future, forest C
estimates in EFDB can be improved through targeted research to fill
critical gaps, reporting of information required by IPCC, and continued
submission of data from scientific publications to the EFDB. Given that
climate change is rapidly impacting the world's forests, timely
reporting of recent estimates will be especially critical to accurate
forest C accounting.

@@

The Global Forest Carbon Database, ForC
(\url{https://forc-db.github.io/}), is the largest collection of
published estimates of forest C stocks, increments, and annual fluxes
\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
includes data ingested from individual publications and relevant
databases, including the Global Reforestation Opportunity Assessment
(GROA) database \citep[database doi:
10.5281/zenodo.3983644]{cook-patton_mapping_2020}, the global soil
respiration database
\citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
For C currently contains 39855 records from 10589 plots in 1535 distinct
geographical areas, along with records of stand age and disturbance
history. As such, ForC is positioned to improve forest C accounting
through the transfer of data to EFDB. The purpose of this publication is
to document that process and provide recommendations for future
improvements.

@@

We added a total of 15 new EFDB-relevant variables to the set of named
and defined variables (Fig. 1), counting each pair of variables with
units in C (ending in \texttt{\_C}) or organic matter (ending in
\texttt{\_OM}) as one. This included eleven increment variables, adding
to only one previously defined increment variable (aboveground biomass
increment, \emph{delta.agb}). These are directly related to C stocks as
previously defined in ForC, with ``\emph{delta.}'' added in front of the
variable name. Further, we added variables capturing the belowground
component of woody mortality (\emph{woody.mortality\_root}) and the
combined aboveground and belowground components of woody mortality
(\emph{woody.mortality}). Although most of these variables currently
lack records in ForC, the structure exists such that records can be
populated over time. Finally, to provide better definition of the
previously existing variable \emph{organic.layer}, which has a nebulous
definition that reflects the varied definitions adopted by original
studies, we added two clearly defined variables: \emph{litter}
(relatively undecomposed plant material/ OL horizon), and
\emph{O.horizon} (entire O-horizon, including OL).

@@

Prior to releasing ForC v4.0, we executed several quality control
measures. First, we implemented a system of continuous integration using
GitHub Actions \citep[\emph{sensu}][]{kim_using_inreview} to run some
automatic checks any time the master data files are updated, including
outlier tests and checks for completeness and naming consistency of
records across data files. Second, to improve information on geographic
coordinates, we created a field to record coordinate precision (Appendix
A), and flagged and reviewed records with suspected low precision
\emph{(Issue \#29){[}https://github.com/forc-db/ForC/issues/229{]}}.
Third, to identify erroneous climate data\ldots{} \emph{(Issue
\#212){[}https://github.com/forc-db/ForC/issues/212{]}}.

Because ForC v4.0 contained known duplicate records, we used R scripts
to remove likely duplicates, as detailed in
\citet{anderson-teixeira_carbon_2021}. Henceforth, we refer to the set
of records with likely duplicates removed as ``independent records''.
All records sent to EFCB were ensured to be original through manual
review, as detailed below.

@@

To transfer complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
records had been checked against the original study and determined to be
complete and correct (\texttt{EFDB.ready} field in
\texttt{ForC\_citations} table), (2) the original study presented values
in tables or text, as opposed to the values having been digitized from
graphs or calculated based on related variables
(\texttt{data.location.within.source} field in
\texttt{ForC\_measurements} table), and (3) the records had not
previously been sent to EFDB. Once converted into EFDB format, the
records were reviewed and then sent to the IPCC's Technical Support Unit
for inclusion in EFDB. Complete records needed to be reviewed by the
EFDB editorial board prior to posting in the database -- a process that
lags behind records transfer and had not yet been completed for all
records sent as of May 05, 2023.

@@

As of May 05, 2023, ForC (v4.0) contained 32693 independent records
(39855 total), 17204 of which were for variables relevant to EFDB (Fig.
1). These records were distributed across all forested continents and
ecozones, with particularly high concentrations in
\emph{{[}ecozone(s){]}} and low concentrations in
\emph{{[}ecozone(s){]}} (Fig. 2). The most widely represented forest
type was \emph{{[}type{]}}, followed by \emph{{[}type{]}} and
\emph{{[}type{]}} (Fig. 3).

@@

As of May 05, 2023, we had reviewed or added 2034 EFDB-relevant records,
1214 records of which were sent to EFDB, and 73 of which have been
reviewed, accepted, and posted (Figs. 2-3, Table 2).
\emph{{[}DETAILS{]}}

>>>>

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned emissions reductions
under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accesible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
accounting if made available through the Emission Factor Database (EFDB)
of the International Panel on Climate Change (IPCC). Here, we develop
and document a process for semi-automated transfer of data from ForC
into the EFDB, assess the data available and transferred to date, and
provide recommendations for improving forest data collection, analysis,
and reporting to improve accounting of forest-sector greenhoouse gas
emissions and removals. We begin by reconciling terminology and mapping
ForC fields into EFDB. This process required some updates to the ForC
database structure, leading to the release of a new version of ForC
(v4.0; described here). As of May 06, 2023, ForC contained
\textasciitilde17204 independent records that would be relevant to EFDB,
1214 of which have been submitted to date. Among the data in ForC, there
is disproportionate representation of biomass (particularly aboveground)
stocks, with far fewer records for dead organic matter and soil C, and
relatively few or no records for net annual increments or C fluxes into
(gains) or out of (losses) the IPCC-defined C pools. Geographic
representation is also quite uneven, with the highest densities of
relevant records in temperate forests, and with relatively scant
representation of tropical forests in Africa and Asia. ForC represents a
diversity of stand ages, although records for young stands are primarily
limited to C stocks, as opposed to net increments of fluxes. This
distribution of records is generally reflected in the subset of records
that have been submitted to EFDB to date. In the future, forest C
estimates in EFDB can be improved through targeted research to fill
critical gaps, reporting of information required by IPCC, and continued
submission of data from scientific publications to the EFDB. Given that
climate change is rapidly impacting the world's forests, timely
reporting of recent estimates will be especially critical to accurate
forest C accounting.

@@

The Global Forest Carbon Database, ForC
(\url{https://forc-db.github.io/}), is the largest collection of
published estimates of forest C stocks, increments, and annual fluxes
\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
includes data ingested from individual publications and relevant
databases, including the Global Reforestation Opportunity Assessment
(GROA) database \citep[database doi:
10.5281/zenodo.3983644]{cook-patton_mapping_2020}, the global soil
respiration database
\citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
As of May 06, 2023, ForC contained 39855 records from 10589 plots in
1535 distinct geographical areas, along with records of stand age and
disturbance history. As such, ForC is positioned to improve forest C
accounting through the transfer of data to EFDB. The purpose of this
publication is to document that process and provide recommendations for
future improvements.

@@

We added a total of 15 new EFDB-relevant variables to the set of named
and defined variables (Fig. 1), counting each pair of variables with
units in C (ending in \texttt{\_C}) or organic matter (ending in
\texttt{\_OM}) as one. The majority of these were increment variables
(n=11), adding to only one previously defined increment variable
(aboveground biomass increment, \emph{delta.agb}). These are directly
related to C stocks as previously defined in ForC, with
``\emph{delta.}'' added in front of the variable name. Further, we added
variables capturing the belowground component of woody mortality
(\emph{woody.mortality\_root}) and the combined aboveground and
belowground components of woody mortality (\emph{woody.mortality}).
Although most of these variables lacked records in ForC as of May 06,
2023, their addition gave the structure such that records can be
populated over time. Finally, to provide better definition of the
previously existing variable \emph{organic.layer}, which has a nebulous
definition that reflects the varied definitions adopted by original
studies, we added two clearly defined variables: \emph{litter}
(relatively undecomposed plant material/ OL horizon), and
\emph{O.horizon} (entire O-horizon, including OL).

@@

Prior to releasing ForC v4.0, we executed several quality control
measures. First, we implemented a system of continuous integration using
GitHub Actions \citep[\emph{sensu}][]{kim_implementing_2022} to run some
automatic checks any time the master data files are updated, including
outlier tests and checks for completeness and naming consistency of
records across data files. Second, to improve information on geographic
coordinates, we created a field to record coordinate precision (Appendix
A), and flagged and reviewed records with suspected low precision.
Third, to identify erroneous climate data, we compared ForC climate
values to those extracted from WorldClim version \textbf{\#\#}
\citep{climwin_ref} based on site coordinates. Records deviating from
WorldClim values by more variable-specific thresholds (\textgreater5°C
for mean annual temperature, \textgreater7.5°C for mean temperatures of
the warmest and coldest months, or \textgreater1 for log(mean annual
precipitation in mm)) were flagged as requiring review prior to use in
analysis or transfer to EFDB.

Because ForC v4.0 contained known duplicate records, we used R scripts
to identify likely duplicates, as detailed in
\citet{anderson-teixeira_carbon_2021}. Henceforth, we refer to the set
of records with likely duplicates removed as ``independent records''.
All records sent to EFDB were ensured to be independent and original
through manual review, as detailed below.

@@

To transfer complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
records had been checked against the original study and determined to be
complete and correct (\texttt{EFDB.ready} field in
\texttt{ForC\_citations} table), (2) the original study presented values
in tables or text, as opposed to the values having been digitized from
graphs or calculated based on related variables
(\texttt{data.location.within.source} field in
\texttt{ForC\_measurements} table), and (3) the records had not
previously been sent to EFDB. Once converted into EFDB format, the
records were reviewed and then sent to the IPCC's Technical Support Unit
for inclusion in EFDB. Complete records needed to be reviewed by the
EFDB editorial board prior to posting in the database -- a process that
lags behind records transfer and had not yet been completed for all
records sent as of May 06, 2023.

@@

As of May 06, 2023, ForC (v4.0) contained 32693 independent records
(39855 total), 17204 of which were for variables relevant to EFDB (Fig.
1). These records were distributed across all forested continents and
ecozones, with particularly high concentrations in
\emph{{[}ecozone(s){]}} and low concentrations in
\emph{{[}ecozone(s){]}} (Fig. 2). The most widely represented forest
type was \emph{{[}type{]}}, followed by \emph{{[}type{]}} and
\emph{{[}type{]}} (Fig. 3).

@@

As of May 06, 2023, we had reviewed or added 2034 EFDB-relevant records,
1214 records of which were sent to EFDB, and 73 of which have been
reviewed, accepted, and posted (Figs. 2-3, Table 2).
\emph{{[}DETAILS{]}}
